# The R *Environment*

R is an environment for statistical modeling and graphics. That it is an
*environment* instead of a *package* is one of the main reasons why R is 
distinct from other statistical programs. A package only gives us a fixed set of
tools. An environment allows us to modify, combine and even create tools to suit
our specific needs.

## Why becoming a useR?

R is gratis and runs on Windows, MacOS, and several Unix platforms. With R you can start with a data set like this:
```{r loading messy flower data}
#| echo: false
flower_df <- read.table("data_files/flower.csv", header = TRUE, sep = ",")
head(flower_df, 5)
```

and, in 8 lines of code or less, make a plot like this:
```{r height by nitrogen boxplots}
#| echo: false
boxplot(
    height ~ nitrogen,
    data = flower_df, 
    col = c("yellow", "blue", "pink"),
    main = "No clear association between height and nitrogen level",
    xlab = "Nitrogen",
    ylab = "Height"
)
```

You can also find many powerful tools to fit statistical models 
(both bayesian and frequentist), including:

- Generalized linear models (including linear regression)
- Survival analysis
- Time series analysis
- Multilevel models (aka hierarchical models, aka Random and Mixed effects 
    models)
- Classification and clustering
- Sample size and power calculations
- Multivariable analysis (e.g., factor analysis, principal component analysis, 
    and structural equations modeling)

Even better, new tools become available in R all the time because, as with
other open source programming languages, everyone can examine and contribute to
R's code. Users constantly publish their own code packages to expand R's
capabilities. As of June 2025, there are over 22 thousand packages to the
Comprehensive R Archive Network (CRAN)[^metacran]. Many of these packages
perform complex statistical routines that are not (and may never be) available
in other statistical software systems.

[^metacran]: r-pkg.org. (n.d.). Retrieved July 3, 2025, from
https://www.r-pkg.org/

In Windows, there are several ways to use R. The standard R graphical
unit interface (GUI) allows you to point and click to do many basic tasks.
Another GUI is [R Commander](https://socialsciences.mcmaster.ca/jfox/Misc/Rcmdr/), developed by 
John Fox at McMaster University. R Commander displays the underlying R
code for each analysis to help the user learn the programming language.
[Tinn-R](http://sourceforge.net/projects/tinn-r/)  is another GUI from Jose
Claudio Faria. 

These GUIs are friendly and easy to grasp if you're a beginner. But to use all
of R's capabilities you will need to do more than point and click. A more
complete way of using R is through an integrated development environment 
(IDE), which, in short, helps you code. The most popular IDE for R is
[RStudio](https://www.posit.com), which organizes the user's screen into
panes that display scripts, objects, graphics, and the R console.

In these notes, we will use RStudio a lot. The goal is for you to start taking full advantage of R's capabilities.

<!--

### Easier to make reproducible research

R's availability and accessibility mean that you can share your work with many. This allows consumers to confirm the
original analysis and to investigate other approaches.

-->


## Why Isn't Everyone a UseR?

Many users of statistics don't use R because they only know how to use one
statistical software, often the one taught in their first statistics course. In
the past, R rarely was this first language, but nowadays more schools are
teaching how to use it.

Some people have used R, but struggled to get comfortable and productive
with it, especially if they had little coding experience. Typing commands
explicitly is more difficult than pointing and clicking. Also, each package has
its own rules to learn. We can find a lot of good help for popular packages
written by professional developers, but not so much for smaller packages written
by other amateur users. Worst of all, some of the messages R displays if we make
a mistake are uninformative, so fixing the problem can be difficult.

Don't get frustrated! You don't have to be an expert programmer to use
R, and the benefits are worth spending some time up front.

## Suggestions for Learning R

- Learn interactively! Retype and experiment with lots of sample code; you won't
    break it. These notes contain many code examples and you can find many more
    online.
- Don't worry about getting errors. Even experienced R users make mistakes
    all the time, and you can learn a lot from error messages.
- Ask other R users for help.
- Some useful links are:
    -   <https://www.r-project.org>: The R Home page, the
        central webpage for the R project. Here you will find
        links for downloading R, downloading additional
        packages for R, and almost everything else that you
        would like to know about the software or the people behind it.
    -   <https://cran.r-project.org/web/views/>: Task views summarize the most
        important packages involved in a subject field or analysis type.
    -   <https://journal.r-project.org>: The R Journal
    -   <https://stats.stackexchange.com>: Cross-Validated
    -   <https://www.r-bloggers.com>
    -   <https://stats.idre.ucla.edu/r/>: Institute for Digital Research and
        Education at UCLA
    -   <https://socialsciences.mcmaster.ca/jfox/>: John Fox's home page
    -   <https://sas-and-r.blogspot.com/>: Examples of code to perform same task
        in SAS and R.

## How to get R

At the [R Project Web Page](https://www.r-project.org) the
most important link is at the left hand side of the screen, under the "Download"
heading. Click on the CRAN link (Comprehensive R Archive
Network), and, after you choose one of the U.S. mirrors, you will be taken to
the page that you will use to download everything R-related.

Once you find the CRAN web page, take the following steps to obtain
R:

1.  Click on "Download R for X" that best describes your operating system       (Linux, OS X, Windows).
2.  When using Windows, click on the "base" subdirectory. This will allow
    you to download the base R packages.
3.  Click the "Download R 4.X.X for Windows" link.
    R is updated quite frequently. At the time of this
    printing, version 4.4.2 is available. Save the `.exe` file somewhere on your
    computer.
4.  Double-click on the `.exe` file once it is downloaded. An installation
    window will appear to guide you through the setup of R in your machine.
5.  Once you finish, you should have an R icon on your
    desktop that gives you a shortcut to the R system.

### How to get RStudio

RStudio is already installed on the lab workstations. The
following information is useful if you need to install
RStudio on another machine. **You must install R before you install RStudio**. Otherwise, RStudio will not work.

Visit <https://posit.co/downloads/> and click on "Download RStudio". Choose the version for your operating system (Linux, OS X, Windows) and download the installer. Then double click on the installer `.exe` file and follow the instructions on the screen to install RStudio.

<!---
### Contributed Packages

What exactly are "additional contributed packages"? R is an
open source software environment, so users are free to explore the code behind
the software and to write their own new code. Statisticians, researchers, and
other users have written additional packages for R that
perform complex analyses (and simple ones!). In order to use these packages and
the functions within them, you first need to download them. Base
R comprises many packages, but odds are that you will
discover an uncommon analysis technique in your research that requires you to
install an additional package that is not included by default.

These additional packages are usually found CRAN. However, there are other
repositories as well (e.g., bioconductor, github).
!-->