# The R *Environment*

**R** is an environment for statistical
modeling and graphics. That it is an *environment* instead of a *package* is one of the main reasons **R** is distinct from other statistical
programs. A package only gives us a fixed set of tools. An environment
allows us to modify, combine and even create tools to suit our specific needs.

## Why becoming a useR?

### Big potential

You can start with a data set that looks like this:
```{r loading messy flower data}
#| echo: false
flower_df <- read.table("data_files/flower.csv", header = TRUE, sep = ",")
head(flower_df, 5)
```

And, in 8 lines of code or less, you can get a plot that looks like this:
```{r height by nitrogen boxplots}
#| echo: false
boxplot(
    height ~ nitrogen,
    data = flower_df, 
    col = c("yellow", "blue", "pink"),
    main = "No clear pattern between height and nitrogen level",
    xlab = "Nitrogen",
    ylab = "Height"
)
```

### Price (\$0) and availability

**R** is free---as in "no money required"---and it runs on
Windows, MacOS, and several Unix platforms.

### Easy and cool graphics

With **R**, you can make basic scatter plots or histograms with one or two lines of code (assuming your data has a nice format). And you can use
the "grammar of graphics" to make incremental changes: add custom colors, change
the size of the plot or its contents, or arrange several graphics in grids.

### Abundant statistical tools

**R** offers many powerful tools to fit statistical models 
(both bayesian and frequentist), including:

- Linear and Generalized Linear Models
- Survival analysis
- Time series analysis
- Random and Mixed Effects Models (aka hierarchical models, aka multilevel
    models)
- Classification and Clustering
- Sample Size and Power Calculations
- Multivariable analysis (Factor Analysis, Principal Component Analysis, Structural Equations Modeling, ...)

### Constant expansion

New tools become available in R all the time. **R** is a
full-fledged programming language that allows users to write their own
functions. Also, all of **R**'s code is open source, which means
that everyone can examine it and contribute to it. Users constantly publish
their own "packages" (ready-to-use tools), which expand **R**'s
base capabilities. As of March 2019, users have contributed over 13,700 packages
to Comprehensive **R** Archive Network (CRAN). Many of these
packages perform complex statistical routines that are not (and may never be)
available in other statistical software systems.

### Flexible interfaces

In Windows, there are several ways to use **R**. The standard 
**R** graphical unit interface (GUI) allows you to point and 
click to do many basic tasks. Another GUI is [**R** 
Commander](https://socialsciences.mcmaster.ca/jfox/Misc/Rcmdr/), developed by 
John Fox at McMaster University. **R** Commander displays the 
underlying **R** code for each analysis to help the user learn
the programming language. [Tinn-**R**](http://sourceforge.net/projects/tinn-r/) 
is another GUI from Jose Claudio Faria.

These GUIs are friendly and easy to grasp if you're a beginner. But to use all of **R**'s capabilities you will need to do more than point and click.

A more complete way of using **R** is through an integrated development environment (IDE), which, in short, helps you code. The most popular IDE for **R** is [**R**Studio](https://www.posit.com), which organizes the
user's screen into panes that display scripts, objects, graphics, and
the **R** console.

In these notes, we will use **R**Studio a lot. The goal is for you to start taking full advantage of **R**'s capabilities.

<!--

### Easier to make reproducible research

**R**'s availability and accessibility mean that you can share your work with many. This allows consumers to confirm the
original analysis and to investigate other approaches.

-->


## Why Isn't Everyone a Use**R**?

Maybe because many users of statistics only learn how to use one statistical software, often the one taught in their first statistics course. In the past, **R** was almost never this first language, but nowadays many schools are teaching how to use it.

And even if people have used **R**, they can take time to get comfortable and 
productive with it, especially if you have little coding experience. Typing
commands explicitly is more difficult than pointing and clicking. Also, each 
package has its own rules to learn. You can find a lot of good help for popular 
packages written by professional developers, but not so much for smaller 
packages written by other common users. Worst of all, some of the messages **R**
displays if you make a mistake are uninformative, so figuring out what went 
wrong can be difficult. 

Don't get frustrated! You don't have to be an expert programmer to use
**R**. The benefits are worth spending some time up front.

## Suggestions for Learning **R**

- Learn interactively! Retype and experiment with lots of sample code; you won't   break it. These notes contain several code examples and you can find many more   online.
- Don't worry about getting errors. Experienced **R** users
  make errors all the time. Besides, you can learn a lot from error
  messages.
- Ask other **R** users for help.
- Some useful links are:
    -   <https://www.r-project.org>: The **R** Home page, the
        central webpage for the **R** project. Here you will find
        links for downloading **R**, downloading additional
        packages for **R**, and almost everything else that you
        would like to know about the software or the people behind it.
    -   <https://cran.r-project.org/web/views/>: Task views summarize the most
        important packages involved in a subject field or analysis type.
    -   <https://journal.r-project.org>: The **R** Journal
    -   <https://stats.stackexchange.com>: Cross-Validated
    -   <https://www.r-bloggers.com>
    -   <https://stats.idre.ucla.edu/r/>: Institute for Digital Research and
        Education at UCLA
    -   <https://socialsciences.mcmaster.ca/jfox/>: John Fox's home page
    -   <https://sas-and-r.blogspot.com/>: Examples of code to perform same task
        in each language

## How to get **R**

**R** is already installed on the lab workstations. The following
information is useful if you need to install **R** on another
machine.

At the [**R** Project Web Page](https://www.r-project.org) the
most important link is at the left hand side of the screen, under the "Download"
heading. Click on the CRAN link (Comprehensive **R** Archive
Network), and, after you choose one of the U.S. mirrors, you will be taken to
the page that you will use to download everything **R**-related.

Once you find the CRAN web page, take the following steps to obtain
**R**:

1.  Click on "Download **R** for XXX" that best describes your
    operating system (Linux, OS X, Windows).
2.  When using Windows, click on the "base" subdirectory. This will allow
    you to download the base **R** packages.
3.  Click the "Download **R** 3.X.X for Windows" link.
    **R** is updated quite frequently. At the time of this
    printing, version 4.3.3 is available. Save the `.exe` file somewhere on your
    computer.
4.  Double-click on the `.exe` file once it has been downloaded. A wizard will
    appear that will guide you through the setup of the **R**
    software on your machine.
5.  Once you are finished, you should have an **R** icon on your
    desktop that gives you a shortcut to the **R** system.
    Double-click on this icon, and you are ready to go!

### How to get RStudio

**R**Studio is already installed on the lab workstations. The
following information is useful if you need to install
**R**Studio on another machine. **You must install R before you install RStudio**. Otherwise, **R**Studio will not work.

Visit <https://www.posit.co> and download the free desktop version of
**R**Studio.

### Contributed Packages

What exactly are "additional contributed packages"? **R** is an
open source software environment, so users are free to explore the code behind
the software and to write their own new code. Statisticians, researchers, and
other users have written additional packages for **R** that
perform complex analyses (and simple ones!). In order to use these packages and
the functions within them, you first need to download them. Base
**R** comprises many packages, but odds are that you will
discover an uncommon analysis technique in your research that requires you to
install an additional package that is not included by default.

These additional packages are usually found CRAN. However, there are other
repositories as well (e.g., bioconductor, github).
