---
title: "Intro to R with RStudio"
author: "Abner Heredia Bustos, CSCAR"
format:
  revealjs:
    incremental: true
---

# Welcome

## About this workshop

+ For novice or inexperienced coders that want to use **R**.
+ From 1 to 5 pm.
+ A few slides for context and extra information.
+ A lot of hands-on coding and live demonstrations.

## What is CSCAR?

+ Full name: Consulting for Statistics, Computing and Analytics Research.
+ A unit of the Office of the Vice President of Research.
+ Guides and trains researchers in data collection, management, and analysis.
+ Also helps researchers to use technical software and advanced computing.
+ Address: The University of Michigan, 3560 Rackham, 915 E. Washington St., Ann Arbor, MI 48109-1070.

## CSCAR is here to help you

+ Free, one-hour consultations with graduate-level statisticians.
+ To request a consultation: email <deskpeople@umich.edu>, or fill [this form](https://docs.google.com/forms/d/e/1FAIpQLSei-twcjFkoobUrVwSQTmSxdKKEc1Ub8w5LHmeIZUmTV1wmIg/viewform?pli=1). Or visit [cscar.research.umich.edu](cscar.research.umich.edu).
+ Walk-in consultations with GSRAs. Self-schedule using [this link](https://calendar.google.com/calendar/u/0/selfsched?sstoken=UUMyTFpCR1RXbmhYfGRlZmF1bHR8ZWNjNGJlMWZlYTA4ZWE5NzYzNmNkNzgyZjUyZDYxNDg). 
+ GSRAs are available Monday through Friday, between 9am and 5pm (we close on Tuesdays between noon and 1pm). 
+ All of our scheduled appointments can be either remote or in-person.

## Who am I?

+ Abner Heredia Bustos, a data science consultant at CSCAR.
+ I want to code with as little time and effort as possible...
+ ...which means learning it well from the beginning.

# Why becoming a useR? 

## R is cool for statistics and graphics

+ **R** is gratis and it runs on Windows, MacOS, and several Unix platforms. 
+ You can start with this:

---

```{r loading messy flower data}
#| echo: false
flower_df <- read.table("data_files/flower.csv", header = TRUE, sep = ",")
head(flower_df, 5)
```

---

and, in 8 lines of code or less, make this:
```{r height by nitrogen boxplots}
#| echo: false
boxplot(
    height ~ nitrogen,
    data = flower_df, 
    col = c("yellow", "blue", "pink"),
    main = "No clear pattern between height and nitrogen level",
    xlab = "Nitrogen",
    ylab = "Height"
)
```

## R is an *environment*, not a package

+ A package is a fixed set of tools. 
+ An environment is for combining, modifying, and creating tools.

## R has plenty of statistical tools and models

+ Generalized linear models (including linear regression).
+ Survival analysis.
+ Time series analysis.
+ Random and Mixed effects models (aka hierarchical models, aka multilevel
    models).
+ Classification and clustering.
+ Sample size and power calculations.
+ Multivariable analysis (e.g., factor analysis, principal component analysis, and structural equations modeling).
+ And more.

## Even more tools and models

+ Users constantly publish their own code packages: more than 13,700 in the Comprehensive **R** Archive Network (CRAN) as of March 2019.
+ Many complex statistical routines are not (and may never be) available in other statistical software.

## Why Isn't Everyone a Use**R**?

+ Some people only use whatever they learned first, which is not always **R**.
+ Other software seems friendlier.
+ Need to learn rules of packages you use.
+ Help pages and error messages may be hard to understand.

## Suggestions for Learning **R**

+ Learn interactively.
+ Don't worry about getting errors.
+ Ask other **R** users for help.

## Some useful links

::: {.nonincremental}

-   <https://www.r-project.org>: Here you will find links for downloading **R**,
    downloading additional packages for **R**, and more.
-   <https://cran.r-project.org/web/views/>: Summaries of important packages         by subject field or analysis type.
-   <https://journal.r-project.org>: The **R** Journal.
-   <https://stats.stackexchange.com>: Cross-Validated.
-   <https://www.r-bloggers.com>
    
:::

## More useful links

::: {.nonincremental}

-   <https://stats.idre.ucla.edu/r/>: Institute for Digital Research and
    Education at UCLA.
-   <https://socialsciences.mcmaster.ca/jfox/>: John Fox's home page.
-   <https://sas-and-r.blogspot.com/>: Examples of code to perform same task
    in SAS and R.
    
:::

# Let's start coding





